Communication complexity: Chapter 3 \\ Distributional communication complexity

### 2019-04-12 CS860-CommunicationComplexity/assignments/Chapter3.tex
Correcting Chapter 2
Corruption bound proofs remain in fixing chapter 3

Consider a partition of the input space into $f$-monochromatic rectangles, $\mathcal R$, such that $|\mathcal R| = \chi(f)$. Then
$$\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|  = \sum_{R\in \mathcal R} \max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right| \geq\sum_{R \in \mathcal{R}} \left| \sum_{(x,y) \in R} \big(M^{\pm}_f\big)_{x,y} \right|= \sum_{R \in \mathcal R} |R|$$
using that each $R$ is $f$-monochromatic to obtain the last equality. Since $\mathcal R$ partitions the input space,
$$\sum_{R \in \mathcal R} |R| = |\calX||\calY|$$
and hence the result holds.
\end{proof}

@@

\begin{proof}
Let $\pi$ be the protocol attaining $D_\epsilon^{\mathrm{unif}}(f)$. Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
$$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
$$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr((x,y) \in R_i)\Pr(\pi(x,y) = f(x,y) \mid(x,y) \in R_i) = \sum_{i=1}^t \frac{|R_i|}{|\calX||\calY|}\frac{|R_i\cap f^{-1}(\pi(x,y))|}{|R_i|}.$$
Symmetrically
$$\Pr[\pi(x,y) \neq f(x,y)] = \frac{1}{|\calX||\calY|} \sum_{i=1}^t|R_i \cap f^{-1}(\pi(x,y)\oplus 1)|.$$
Thus we have
\begin{align*}1-2\epsilon &= \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)] \\
&= \frac{1}{|\calX||\calY|} \sum_{i=1}^t|R_i\cap f^{-1}(\pi(x,y))| -|R_i \cap f^{-1}(\pi(x,y)\oplus 1)| \\
&\leq \frac{1}{|\calX||\calY|} \sum_{i=1}^t  \left| \sum_{(x,y) \in R_i} \big(M^{\pm}_f\big)_{x,y} \right|\\
&\leq t\cdot\discu(f).\end{align*}
Since $\log_2(t) = D_\epsilon^{\mathrm{unif}}(f)$ this completes the proof.
\end{proof}

@@

\begin{proof}
For convenience, define $M:= M^{\pm}_\IP$ and let $\mathbf{1}_S$ denote characteristic vector of a set $S\subseteq [n]$. Now consider an entry $x,y$ of $M^TM$:
$$M^TM_{x,y} = \sum_{z\in\{0,1\}^n} (-1)^{x\cdot z}(-1)^{y\cdot z} = \sum_{z \in \{0,1\}^n}(-1)^z\cdot(x+y) = \begin{cases}
(-1)2^{n-1} + (1)2^{n-1}, &\text{if $x\neq y$} \\
\sum_{z \in \{0,1\}^n}1, &\text{otherwise}.
\end{cases}$$
Hence $$M^TM_{x,y} = \begin{cases}
0, &\text{if $x\neq y$} \\
2^n, &\text{otherwise}
\end{cases}$$
i.e. $M^TM = 2^nI$.

@@

\begin{proof}
Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$. Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
$$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
$$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr[(x,y) \in R_i \text{ and }\pi(x,y) = f(x,y) ] = \sum_{i=1}^t \mu(R_i \cap f^{-1}(\pi(x,y))).$$
Symmetrically
$$\Pr[\pi(x,y) \neq f(x,y)] = \sum_{i=1}^t\mu(R_i \cap f^{-1}(\pi(x,y)\oplus 1)).$$
Thus we have
\begin{align*}1-2\epsilon &= \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)] \\
&=\sum_{i=1}^t|\mu(R_i\cap f^{-1}(\pi(x,y))) -\mu(R_i \cap f^{-1}(\pi(x,y)\oplus 1))| \\
&\leq \sum_{i=1}^t \disc_\mu(f)\\
&= t\cdot\disc_\mu(f).\end{align*}
Since $\log_2(t) = D_\epsilon^{\mu}(f)$ this completes the proof.
\end{proof}

@@

\begin{proof}
I'm not really sure what the right distribution to take is, but intuitively I want to distribute the probability mass evenly between $(x,y)$ pairs with $x\neq y$ and pairs with $x=y$ so Alice and Bob have to exchange some bits to do better than guessing. From the matrix of this function I want half the probability mass uniformly distributed on the diagonal, and half uniformly distributed on all off-diagonal entries.
\end{proof}

@@

>>>>

Consider a partition of the input space into $f$-monochromatic rectangles, $\mathcal R$, such that $|\mathcal R| = \chi(f)$. Then
\begin{align*}\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|  &= \sum_{R\in \mathcal R} \max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|\\
&\geq\sum_{R \in \mathcal{R}} \left| \sum_{(x,y) \in R} \big(M^{\pm}_f\big)_{x,y} \right|\\
&= \sum_{R \in \mathcal R} |R|
\end{align*}
using that each $R$ is $f$-monochromatic to obtain the last equality. Since $\mathcal R$ partitions the input space,
$$\sum_{R \in \mathcal R} |R| = |\calX||\calY|$$
and hence the result holds.
\end{proof}

@@

\begin{proof}
Let $\pi$ be the protocol attaining $D_\epsilon^{\mathrm{unif}}(f)$. Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
$$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
\begin{align*}\Pr[\pi(x,y) = f(x,y)] &= \sum_{i=1}^t \Pr((x,y) \in R_i)\Pr(\pi(x,y) \\
&= f(x,y) \mid(x,y) \in R_i)\\
&= \sum_{i=1}^t \frac{|R_i|}{|\calX||\calY|}\frac{|R_i\cap f^{-1}(\pi(x,y))|}{|R_i|}.
\end{align*}
Symmetrically
$$\Pr[\pi(x,y) \neq f(x,y)] = \frac{1}{|\calX||\calY|} \sum_{i=1}^t|R_i \cap f^{-1}(\pi(x,y)\oplus 1)|.$$
Thus we have
\begin{align*}1-2\epsilon &= \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)] \\
&= \frac{1}{|\calX||\calY|} \sum_{i=1}^t|R_i\cap f^{-1}(\pi(x,y))| -|R_i \cap f^{-1}(\pi(x,y)\oplus 1)| \\
&\leq \frac{1}{|\calX||\calY|} \sum_{i=1}^t  \left| \sum_{(x,y) \in R_i} \big(M^{\pm}_f\big)_{x,y} \right|\\
&\leq t\cdot\discu(f).\end{align*}
Since $\log_2(t) = D_\epsilon^{\mathrm{unif}}(f)$ this completes the proof.
\end{proof}

@@

\begin{proof}
For convenience, define $M:= M^{\pm}_\IP$ and let $\mathbf{1}_S$ denote characteristic vector of a set $S\subseteq [n]$. Now consider an entry $x,y$ of $M^TM$:
$$M^TM_{x,y} = \sum_{z\in\{0,1\}^n} (-1)^{x\cdot z}(-1)^{y\cdot z} = \sum_{z \in \{0,1\}^n}(-1)^{z\cdot(x+y)} = \begin{cases}
(-1)2^{n-1} + (1)2^{n-1}, &\text{if $x\neq y$} \\
\sum_{z \in \{0,1\}^n}1, &\text{otherwise}.
\end{cases}$$
Hence $$M^TM_{x,y} = \begin{cases}
0, &\text{if $x\neq y$} \\
2^n, &\text{otherwise}
\end{cases}$$
i.e. $M^TM = 2^nI$.

@@

\begin{proof}
Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$.

Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
$$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
$$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr[(x,y) \in R_i \text{ and }\pi(x,y) = f(x,y) ] = \sum_{i=1}^t \mu(R_i \cap f^{-1}(\pi(x,y))).$$
Symmetrically
$$\Pr[\pi(x,y) \neq f(x,y)] = \sum_{i=1}^t\mu(R_i \cap f^{-1}(\pi(x,y)\oplus 1)).$$
Thus we have
\begin{align*}1-2\epsilon &= \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)] \\
&=\sum_{i=1}^t|\mu(R_i\cap f^{-1}(\pi(x,y))) -\mu(R_i \cap f^{-1}(\pi(x,y)\oplus 1))| \\
&\leq \sum_{i=1}^t \disc_\mu(f)\\
&= t\cdot\disc_\mu(f).\end{align*}
Since $\log_2(t) = D_\epsilon^{\mu}(f)$ this completes the proof.
\end{proof}

@@

\begin{proof}
    We define $\mu : \{0,1\}^n  \times \{0,1\}^n \rightarrow [0,1]$ as follows. For any $(x,y) \in \{0,1\}^n\times \{0,1\}^n$ let
    $$ \mu(x,y) = \begin{cases}
        \frac{1}{2^{n+1}}, &\text{if } x=y \\
        \frac{1}{2^{n+1}(2^n-1)}, &\text{otherwise}.
    \end{cases}.$$
    To see that $\mu$ is a probability distribution observe that
    \begin{align*}\sum_{x,y \in \{0,1\}^n\times \{0,1\}^n} \mu(x,y) &= \sum_{x\in\{0,1\}^n}\mu(x,x) + \sum_{x\neq y \in \{0,1\}^n\times \{0,1\}^n} \mu(x,y)\\
     &= 2^n \frac{1}{2^{n+1}} + (2^{2n}-2^n)\frac{1}{2^{n+1}(2^n-1)}\\
    &= \frac{1}{2} + \frac{1}{2} = 1.
    \end{align*}
    Let $R = A\times B \subseteq \{0,1\}^n\times\{0,1\}^n$ be a rectangle such that
    $$\left|\mu(R\cap \Eq^{-1}(0)) - \mu(R\cap \Eq^{-1}(1))\right| = \disc_\mu(\Eq).$$
    Choose such $R$ so that $|R|$ is minimized. We proceed by case distinction.

    First suppose that $\mu(R\cap \Eq^{-1}(1)) \geq \mu(R\cap \Eq^{-1}(0)).$ If $(x,x), (y,y) \in R\cap \Eq^{-1}(1)$ then $(x,y), (y,x) \in R\cap \Eq^{-1}(0)$. Thus we have that 
    \begin{align*}&\mu(R\cap \Eq^{-1}(1)) - \mu(R\cap \Eq^{-1}(0))\\ 
        &\leq \left|R\cap \Eq^{-1}(1)\right|(\frac{1}{2^{n+1}})-\left(\left|R\cap \Eq^{-1}(1)\right|^2-\left|R\cap \Eq^{-1}(1)\right|\right)(\frac{1}{2^{n+1}(2^n-1)}).
    \end{align*}
    This upperbound is maximized when $|R\cap \Eq^{-1}(1)|=2^{n-1}$. Therefore in this case $\disc_\mu(\Eq) \leq \frac{1}{8} + o(1).$

    Now consider the other case, where $\mu(R\cap \Eq^{-1}(0)) > \mu(R\cap \Eq^{-1}(1)).$ Suppose there exists $x \in \{0,1\}^n$ such that $(x,x) \in R.$ Consider $R' = A\times B\backslash\{x\}.$ Observe that
    \begin{align*}
        &\left|\mu(R'\cap \Eq^{-1}(0)) - \mu(R'\cap \Eq^{-1}(1))\right|\\ & = \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1))  + \mu(x,x)- \sum_{y\in A\backslash\{x\}} \mu(x,y)\\
        &= \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1)) + \frac{1}{2^{n+1}} - |A\backslash\{x\}|\frac{1}{2^{n+1}(2^n-1)} \\
        &\geq \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1)) + \frac{1}{2^{n+1}} - \frac{1}{2^{n+1}} \\
        &= \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1))\\
        &= \disc_\mu(\Eq).
    \end{align*}
    This contradicts that $|R|$ is minimized. Therefore $A\cap B = \emptyset$. Thus we have
    \begin{align*}
        \mu(R\cap \Eq^{-1}(0))-\mu(R\cap \Eq^{-1}(1)) &\leq \mu(R\cap \Eq^{-1}(0))\\
        &= |R\cap \Eq^{-1}(0)|\frac{1}{2^{n+1}(2^n-1)} \\
        &\leq |A|(2^n-|A|)\frac{1}{2^{n+1}(2^n-1)}\\
        &\leq \frac{2^{n}(2^{n/2}-1)}{2^{n+1}(2^n-1)}\\
        &= O\left(\frac{1}{2^{n/2}}\right).
    \end{align*}

    Hence we have shown that $\disc_\mu(\Eq) \leq \frac{1}{8} + o(1) \leq \frac{1}{8}\left(\frac{2^n}{2^n-1}\right)$.
\end{proof}

@@

\begin{proof}
For $r \in \{0,1\}^n$ let  $\pi_r$ be the deterministic protocol where 
$$\pi_r(x,y) = \begin{cases} 1, &\text{if } (x\cdot r)  = (y\cdot r) \\ 0, &\text{otherwise}\end{cases}.$$
This protocol has communication cost $O(1)$ since Alice can send Bob $x\cdot r$ in $1$ bit, and Bob can send Alice $y \cdot r$ in $1$ bit.

Consider a uniformly random $r \in \{0,1\}$. If $x=y$ then $\pi_r(x,y) = 1$. If $x\neq y$ then $\pi_r(x,y) =1$ with probability $\frac{1}{4}$. Let $X_r(x,y)$ be the indicator random variable indicating if $\pi_r(x,y)$ fails on input $(x,y)$. Thus we see that for any $(x,y)$,
$$\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
Thus we have
$$\mathbb{E}_{(x,y)\sim \mu}\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
Rearranning order of expectation yields
$$\mathbb{E}_{r}\mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
And thus there exists an $r \in \{0,1\}$ such that
$$\Pr_\mu(\pi_r \text{ fails}) \leq \mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
\end{proof}
